Msprime 1.0: an efficient and extensible coalescent simulation framework

### 2021-08-05 paper.tex

Crossover recombination is implemented in \msprime\ using Hudson's algorithm, which
works backwards in time, generating common ancestor and recombination
events and tracking their effects on segments of ancestral
material inherited from the
sample~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
Common ancestor events merge the ancestral material of two lineages,
and result in coalescences in the marginal trees when ancestral
segments overlap.
Recombination events split the ancestral material for some lineage
at a given breakpoint, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
us to simulate recombination hotspots and patterns
of recombination from empirical maps.
The implementation of recombination in \msprime\ is extensively validated
against analytical results~\citep{hudson1983properties,kaplan1985use}
and simulations by \ms, \texttt{msHot} and \SLiM.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{figures/ancestry-perf}
\end{center}
\caption{\label{fig-ancestry-perf}
    Running time for \msprime\
    for (A) small and (B) larger simulations on an Intel i7-6600U CPU.
    Each point is the run time of one simulation,
    for various values of effective population size ($N_e$),
    genome length in Morgans ($L$), and number of samples ($n$).
    Run time scales quadratically with the product of $N_e$ and $L$,
    shown on the horizontal axis. For example, (A) shows that
    1,000 samples of 1 Morgan-length chromosomes
    from a population of $N_e=2,000$ diploids
    would take about 2 seconds, and (equivalently) that the same number of
    0.01 Morgan segments with $N_e= 200,000$ would take the same time.
    Since recombination rate in these simulations was $10^{-8}$,
    $L$ is the number of base pairs divided by $10^8$.
    The black lines are quadratic fits separately in each panel
    and sample size. Also shown are the approximate values of
    $N_e L$ for chromosome 1 in three species, using values from
    the \stdpopsim\ catalogue~\citep{adrion2020community}.
}
\end{figure}

@@

In human-like parameter regimes and for large sample sizes,
\msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the population scaled recombination rate
$\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
This is because Hudson's algorithm tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments.
As mentioned above, common ancestor events in which the ancestral material
of two lineages is merged only result in coalescences in the marginal trees
if their ancestral segments overlap. If there is no overlap, the merged
segments represent an ancestral chromosome that is a genetic ancestor
of the two linages, but not the most recent common genetic ancestor
at any location along the genome.
When this happens, the merged lineage carries ``trapped'' genetic material
that is not ancestral to any samples, but where recombinations can still occur.
The SMC approximations work by disallowing common ancestor events
that generate trapped material, greatly simplifying the process.
However, this also removes subtle long-range
correlations in the trees since there are many ways in which ancestry
segments can merge without overlapping.

For large $\rho$, recombination events in trapped ancestral material will
dominate, and so we can use this as a proxy for the overall number
of events in Hudson's algorithm.
\citet[Eq.~5.10]{hein2004gene} gave
\begin{equation}\label{eqn-hsw-bound}
\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2
\end{equation}
as an upper bound
on the number of recombination events within trapped ancestral material
in Hudson's algorithm for $n$ samples.
% Since the sum is well approximated by $\log{n}$,
% a natural conjecture is that the time complexity of Hudson's
% algorithm is $O(\rho^2 \log^2 n)$.
Fig.~\ref{fig-ancestry-perf} shows the observed run time for
simulations with a variety of population size, chromosome
length and sample sizes, and demonstrates that Eq.~\eqref{eqn-hsw-bound}
correctly predicts the quadratic dependence on $\rho$,
as previously conjectured~\citep[Fig.~2]{kelleher2016efficient}.
We also see that the dependence on $n$ is quite weak,
since increasing sample size 100-fold
only increases run time by a factor of 2 or so. However, the
$\log^2{n}$ factor implied by Eq.~\eqref{eqn-hsw-bound}
(the sum is a Harmonic number, and approximated by $\log{n}$)
is not well supported by the data except possibly at very
large values of $\rho$, and so it seems a tighter bound
than Eq.~\eqref{eqn-hsw-bound} would
be required to accurately predict simulation time for a
given $\rho$ and $n$.

Fig.~\ref{fig-ancestry-perf} is a useful yardstick, allowing us
to predict how long simulations should take for a wide range of
species. Taking a typical chromosome to be 1 Morgan in length,
these plots show, roughly, that simulating chromosome-length
samples from a population of thousands of individuals takes seconds,
while samples from a population of tens of thousands takes minutes.
Simulating whole chromosomes for many species is very fast,
with 1000 samples of chromosome 1 for
\emph{Arabidopsis thaliana} taking less than a second, and a few
minutes for dogs and humans. However, the dependence on $\rho$
\emph{is} quadratic, and if $\rho$ is sufficiently large simulations
may not be feasible. For example, for
\emph{Drosophila melanogaster} chromosome 2L, with
$\rho\approx 3.9\times10^6$, Fig.~\ref{fig-ancestry-perf}
predicts that simulation would require around 177 hours. For such
large values of $\rho$  we recommend
users consider approximate simulations. Since \msprime\ does not currently
have efficient implementations of approximate coalescent with recombination
models, we recommend using SMC based methods such as \scrm\,
particularly for moderate sample sizes.
In practice, to predict the running time of a given
simulation in \msprime, we recommend that users
measure run time in a series of simulations with short genome lengths and
the desired sample size,
% possibly varying other parameters to check for interactions,
% JK: why? ^
% I'm worried people will take this as meaning "run every possible
% combination of parameters". I think the genome length should be basically
% independent of everything else, right?
and then predict run time by fitting a quadratic curve to genome length
as in Fig.~\ref{fig-ancestry-perf}.
It is important to note that the quadratic curves in the two
panels of Fig.~\ref{fig-ancestry-perf} are different, and
predicting the run times of days-long simulations
using the timing of minutes-long runs is unlikely to be very accurate.

What about simulations with changing population size?
To understand how run time depends on demography
it helps to consider why run time is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid and inconsequential events that
lead to the quadratic run time.
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of run time we can estimate when this occurs.
Suppose that the maximum number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
Since the maximum number of lineages is reached
very quickly~\citep{nelson2020accounting},
many (roughly half) of these lineages have at
least two distant segments of ancestral material
separated by a long segment of genome, implying that $\ell \approx L / 4$.
Then, since the rates of recombination and common ancestor events
are equal the time the maximum number of lineages is present,
we see that $M$ is roughly equal to $L N_e$.
The number of lineages then decreases gradually from this maximum
on the coalescent time scale, and therefore over roughly $2 N_e$ generations.
Since the total rate of events when the maximum number of lineages is
present is roughly $L^2 N_e / 4$,
then the total number of events is proportional to $(L N_e)^2$---i.e.,
proportional to $\rho^2$.

What does this tell us about time-varying population sizes?
The argument above implies that the work is spread out relatively
evenly on the coalescent time scale.
% JK: I just found this confusing tbh.
% in other words,
% more events happen in the time period from 0 to $N_e/2$ generations
% than from $N_e/2$ to $N_e$ generations,
% but the run time is not dominated by the more recent time period.
Suppose that population size today is $N_1$,
while longer ago than $T$ generations ago it was $N_2$:
does the run time depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
then run time will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the run time will depend critically on $T$---in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.

@@

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{figures/sweeps-perf}
\end{center}
\caption{\label{fig-selection-perf}Comparison of the run time for
\msprime\ and \discoal\ for simulating selective sweeps.
}
\end{figure}

>>>>

Crossover recombination is implemented in \msprime\ using Hudson's algorithm, which
works backwards in time, tracking the
effects of common ancestor and recombination
events~\citep{hudson1983properties,hudson1990gene,kelleher2016efficient}.
Common ancestor events combine the ancestral material of two lineages, and may
result in coalescences in the marginal trees. Recombination events
split the ancestral material for some lineage, creating two independent
lineages. Using the appropriate data structures~\citep{kelleher2016efficient},
this process is much more efficient to simulate than either
methods based on the Ancestral Recombination Graph
formulation~\citep{griffiths1991two,griffiths1997ancestral}
or the left-to-right approach~\citep{wiuf1999recombination,wiuf1999ancestry}.
In \msprime\ 1.0 recombination rates can vary along a chromosome, allowing
us to simulate recombination hotspots and patterns
of recombination from empirical maps.
The implementation of recombination in \msprime\ is extensively validated
against analytical results~\citep{hudson1983properties,kaplan1985use}
and simulations by \ms.

@@

\begin{figure}
\begin{center}
    \includegraphics[width=\textwidth]{figures/ancestry-perf}
\end{center}
\caption{\label{fig-ancestry-perf}
    Running time for \msprime\
    for (A) small and (B) larger simulations on an Intel i7-6600U CPU.
    Each point is the runtime of one simulation,
    for various values of effective population size ($N_e$),
    genome length in Morgans ($L$), and number of samples ($n$).
    Runtime scales quadratically with the product of $N_e$ and $L$,
    shown on the horizontal axis---for instance, figure A shows that
    1,000 samples of 1 Morgan-length chromosomes
    from a population of $N_e=2,000$ diploids
    would take about 2 seconds, and that the same number of
    0.01 Morgan segments in a population of size 200,000
    would take about the same time.
    Since recombination rate in these simulations was $10^{-8}$,
    $L$ is the number of base pairs divided by $10^8$.
    The black lines are quadratic fits separately in each panel
    and sample size. Also shown are the approximate values of
    $\rho$ for chromosome 1 in three species, using values from
    the \stdpopsim\ catalog~\citep{adrion2020community}.
}
\end{figure}

@@

In human-like parameter regimes and for large sample sizes,
\msprime's implementation of the
exact coalescent with recombination comprehensively outperforms
all other simulators, including those based on SMC
approximations~\citep{kelleher2016efficient}. However, it is important
to note that although the implementation of Hudson's algorithm is
very efficient, it is still quadratic in the scaled recombination rate
$\rho = 4 N_e L$, where $L$ is the length of the genome in units of recombination distance.
This is because Hudson's algorithms tracks recombinations not only
in segments ancestral to the sample, but also between ancestral segments.
In other words, lineages may ``coalesce'' even if the pieces of ancestral material they carry
do not overlap---representing an ancestral chromosome
that is a genetic ancestor to each of two linages,
but not the most recent common genetic ancestor at any location along the genome.
When this happens, the resulting merged lineage carries ``trapped'' genetic material
that is not ancestral to any samples;
recombinations in the trapped material will cause the two segments of ancestry
to split apart into separate lineages again.
For large $\rho$, recombination events
in this trapped material will dominate.
\citet[Eq.~5.10]{hein2004gene} gave
$\rho (\rho + 1) \left( \sum_{i=1}^{n-1} \frac{1}{i} \right)^2$
as an upper bound
on the number of recombination events within trapped ancestral material
in Hudson's algorithm with $n$ samples.
Since the sum is well approximated by $\log{n}$,
a natural conjecture is that the time complexity of Hudson's
algorithm is $O(\rho^2 \log^2 n)$.
This turns out to have the correct dependence on $\rho$---although
caution is needed; see below---but the dependence on $n$ is less accurate.

Figure~\ref{fig-ancestry-perf} gives a rough idea of the runtime of
a given simulation of a constant size population,
and verifies that runtime scales quadratically with $\rho$,
and depends much less strongly on the number of samples.
Taking a typical chromosome to be 1 Morgan in length,
these plots show, roughly, that simulating chromosome-length
samples from a population of thousands of individuals takes seconds,
while samples from a population of tens of thousands takes minutes.
However, it is important to note that the quadratic curves in the two
panels are different---predicting the runtimes of days-long simulations
using the timing of minutes-long runs is unlikely to be very accurate.
Further experiments show that adding a $\log^2 n$ factor
to predict runtimes across different sample sizes does not do well
except possibly at very large values of $\rho$.

% Although \citet{hein2004gene} refer to the bound in \eqref{eqn-hsw-bound}
% as ``crude'', it predicts the scaling of run time quite well for large genomes
% Figure~\ref{fig-expected-perf} shows the observed runtime
% (from the same simulations as Figure~\ref{fig-ancestry-perf})
% plotted against equation~\eqref{eqn-hsw-bound},
% showing that for values of $\rho$ above $4 \times 10^4$ or so
% (i.e., chromosome-length simulations in populations above $10^4$),
% the ratio of observed time to the value from \eqref{eqn-hsw-bound} is constant.
% This implies the runtime for a large sample
% can be predicted from the runtime for a small sample.
% For smaller values of $\rho$, however, the prediction
% severely underestimates runtime.

What about simulations with changing population size?
To understand how runtime depends on demography
it helps to consider why runtime is quadratic in $\rho$.
At any point in time, \msprime\ must keep track of some number of lineages,
each of which contains some number of chunks of genetic material.
Common ancestor events reduce the number of lineages,
and recombination events increase their number.
However, with long genomes
only a small fraction of the common ancestor events
will involve overlapping segments of ancestry,
and lead to coalescence in the marginal trees.
Such disjoint segments are often far apart (on average, about distance $L/2$),
and so recombine apart again immediately;
it is these large numbers of rapid yet inconsequential events that
lead to the quadratic runtime.
\plrcomment{(this is discussed also in the ARG section; check for overlap)}
The maximum number of lineages occurs when
the increase and decrease in numbers of lineages due to
common ancestor and recombination events balance out;
to get an idea of runtime we can estimate when this occurs.
Suppose that the maximal number of lineages is $M$;
at this time the rate of common ancestor events is $M (M-1) / (4 N_e)$
and the total rate of recombination is $M \ell$,
where $\ell$ is the mean length of genome carried by each lineage
(including ``trapped'' non-ancestral material).
It is known that the maximum number of lineages is reached
very quickly~\citep{nelson2020accounting},
and at this maximum, recombination and coalescence rates are equal.
Therefore, many (roughly half) of these lineages have at
least two distant segments of ancestral material
separated by a long segment of genome, implying that $\ell \approx L / 4$.
This implies that $M$ is roughly $L N_e$.
The number of linages decreases gradually,
on the coalescent time scale---so, over roughly $2 N_e$ generations.
Since the total rate of events when there are the maximum number of lineages
happens is roughly $L^2 N_e / 4$,
then the total number of events is proportional to $(L N_e)^2$---i.e.,
proportional to $\rho^2$.

What does this tell us about time-varying population sizes?
The argument above implies that the work is spread out relatively
evenly on the coalescent time scale: in other words,
more events happen in the time period from 0 to $N_e/2$ generations
than from $N_e/2$ to $N_e$ generations,
but the runtime is not dominated by the more recent time period.
Suppose that population size today is $N_1$,
while longer ago than $T$ generations ago it was $N_2$:
does the runtime depend more on $4 N_1 L$ or $4 N_2 L$?
The answer depends on how $T$ compares to $N_1$: if $T/N_1$ is large,
then runtime will be similar to a population of size $N_1$;
while if $T/N_1$ is small, it will be similar to a population of size $N_2$.
For instance, in many agricultural species $N_1 \propto 100$, while $N_2 \propto 10^5$,
and the runtime will depend critically on $T$---in other words,
simulation will be quick in a species with a strong domestication bottleneck,
and slow otherwise.

In practice, we recommend that users estimate runtime by
measuring runtime in a series of simulations with short genome lengths,
possibly varying other parameters to check for interactions,
and then predict runtime by fitting a quadratic curve to genome length.
In these experiments, sample sizes should be roughly
those desired---which is not a problem in practice because of
the weak dependency on sample size
---notice that in Figure~\ref{fig-ancestry-perf} increasing sample size 100-fold
only increases runtime by a factor of 2 or so.

@@

\begin{figure}
\begin{center}
\includegraphics[width=6cm]{figures/sweeps-perf}
\end{center}
\caption{\label{fig-selection-perf}Comparison of the runtime for
\msprime\ and \discoal\ for simulating selective sweeps.
}
\end{figure}
